Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-06 doc/invt.tex
add invtpar.h

\documentclass[reprint]{revtex4-1}
\usepackage{amsmath}
\begin{document}

@@

In the WL algorithm, $v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
Note that the updating scheme only applies
to the $v_i$ at the current bin.
%The values of $v_j(t)$ at $j \ne i(t)$
%are kept unchanged.
%
One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p^*_i$.
%
However, the updating scheme
makes $v_i(t)$ a fluctuating quantity,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

@@

The total error of the bias potentials can be written in
terms of $x_i$,
\begin{equation}
\mathcal E
=
\sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
=
\sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle,
\label{eq:error_sum}
\end{equation}
where
$\bar v = \sum_{i = 1}^n p_i v_i$.
%
Our aim is to find the $\alpha(t)$
that minimizes Eq. \eqref{eq:error_sum}.
%
To do so, we shall first consider
a simpler one-variable problem.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eq. \eqref{eq:w_balance},
we can show that
Eqs. \eqref{eq:dvbardt} and \eqref{eq:px_sum}
remain true.
%
%Thus, for Eq. \eqref{eq:x_def}, we have
%
$$
\begin{aligned}
\dot x_i
&= \alpha(t) \sum_{j=1}^n w_{ij}
\left( \frac{ h_j } { p_j }  - 1 \right)
\\
&=
-\alpha(t) \sum_{j = 1}^n
w_{ij} \left[ x_j(t) - \frac{\zeta_j (t)}{p_j} \right],
\end{aligned}
$$
where
we have expanded the right-hand side
in linear terms of $x_j$ in the second step.

@@

A more direct, and often more accurate, approach
is to assume the functional form
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
and find the optimal choice of $\lambda$.
%
Using Eqs. \eqref{eq:kappa_perfect}, \eqref{eq:alpha_trial}
in Eq. \eqref{eq:y2_sum}, we get
%
\begin{equation}
\mathcal E
=
\int_0^t
\sum_{i=0}^n \dot u_i^2(q(t')) \, dt'
\stackrel{t \to \infty}
{=\joinrel=\joinrel=}
\frac{1}{t}
\sum_{i = 2}^n
\frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) },
\label{eq:y2_sum_perfect_invt}
\end{equation}
%
and the optimal $\lambda$
can be found from the smallest root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\end{equation}

@@

%\section{Numerical results}

%Here

>>>>

\documentclass[reprint]{revtex4-1}
\usepackage{amsmath}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

\definecolor{DarkBlue}{RGB}{0,0,64}
\definecolor{DarkBrown}{RGB}{64,20,10}
\definecolor{DarkGreen}{RGB}{0,64,0}
\definecolor{DarkPurple}{RGB}{64,0,42}
% annotation macros
\newcommand{\repl}[2]{{\color{gray} [#1] }{\color{blue} #2}}
\newcommand{\add}[1]{{\color{blue} #1}}
\newcommand{\del}[1]{{\color{gray} [#1]}}
\newcommand{\note}[1]{{\color{DarkGreen}\footnotesize \textsc{Note.} #1}}
\newcommand{\answer}[1]{{\color{DarkBlue}\footnotesize \textsc{Answer.} #1}}
\newcommand{\summary}[1]{{\color{DarkPurple}\footnotesize \textsc{Summary.} #1}}

\begin{document}

@@

In the WL algorithm, $v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
Note that the updating scheme only applies
to the $v_i$ at the current bin.
%The values of $v_j(t)$ at $j \ne i(t)$
%are kept unchanged.
%
One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$.
%
However, the updating scheme
makes $v_i(t)$ a fluctuating quantity,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

@@

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
\mathcal E
=
\sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
=
\sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle,
\label{eq:error_sum}
\end{equation}
where
$\bar v = \sum_{i = 1}^n p_i v_i$.
%
Our aim is to find the $\alpha(t)$
that minimizes Eq. \eqref{eq:error_sum}.
%
To do so, we shall first consider
a simpler one-variable problem.

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eq. \eqref{eq:w_balance},
we can show that
Eqs. \eqref{eq:dvbardt} and \eqref{eq:px_sum}
remain true.
%
%Thus, for Eq. \eqref{eq:x_def}, we have
%
$$
\begin{aligned}
\dot x_i
&= \alpha(t) \sum_{j=1}^n w_{ij}
\left( \frac{ h_j } { p_j }  - 1 \right)
\\
&=
-\alpha(t) \sum_{j = 1}^n
w_{ij} \left[ x_j(t) - \frac{\zeta_j (t)}{p_j} \right],
\end{aligned}
$$
where
we have expanded the right-hand side
in linear terms of $x_j$.

@@

A more direct, and often more accurate, approach
is to assume the functional form
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
and find the optimal value of $\lambda$.
%
Using Eqs. \eqref{eq:kappa_perfect}, \eqref{eq:alpha_trial}
in Eq. \eqref{eq:y2_sum}, we get
%
\begin{equation}
\mathcal E
=
\int_0^t
\sum_{i=0}^n \dot u_i^2(q(t')) \, dt'
\stackrel{t \to \infty}
{=\joinrel=\joinrel=}
\frac{1}{t}
\sum_{i = 2}^n
\frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) },
\label{eq:y2_sum_perfect_invt}
\end{equation}
%
and the optimal $\lambda$
can be found from the smallest root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\end{equation}

@@

\subsection{Example}

As an example, consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij} = m_{|i-j|} + m_{i+j-1} + m_{2n+1-i-j},
  \label{eq:w_band}
\end{equation}
%
where $m_0, \dots, m_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
m_0 + 2 \, m_1 + \cdots + 2 \, m_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $m_l = 0$ for $l > b$.

This matrix is clearly symmetric, and hence
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution $p_i \equiv 1/n$.
%
It admits $n$ eigenvectors,
$\pmb\theta^{(1)}, \dots, \pmb\theta^{(n)}$,
given by
%
\begin{equation}
\theta^{(k)}_i
= \cos \frac{ (2 \, i - 1) k \pi}{2 \, n}.
\label{eq:wband_eigenvector}
\end{equation}
%
with eigenvalues
\note{To show Eqs.
  \eqref{eq:wband_eigenvector} and \eqref{eq:wband_eigenvalue},
  we first define $m_l = 0$, for $l \le 0$, then
$$
\begin{aligned}
&\sum_{i = 1}^n
  \theta^{(k)}_i
  \, w_{ij}
=
- \theta^{(k)}_j \, m_0
\\
&
+\sum_{i=1}^n
  \theta^{(k)}_i
  \left(
    m_{i - j} + m_{j - i} + m_{i + j - 1} + m_{2n+1-i-j}
  \right)
.
\end{aligned}
$$

Now
$$
\begin{aligned}
  &
  \sum_{i=1}^n
  \theta^{(k)}_i \,
  \left(
    m_{i - j} + m_{2n+1-i-j}
  \right)
  \\
  &=
  \sum_{l=1-j}^{n-j} \theta^{(k)}_{j+l} \, m_l
  +
  \sum_{l=n-j+1}^{2n-j} \theta^{(k)}_{2n+1-j-l} \, m_l
  \\
  &=
  \sum_{l=1-j}^{2n-j} \theta^{(k)}_{j+l} \, m_l
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j+l-\frac12)k\pi}{n},
\end{aligned}
$$
where we have used
$$
\theta^{(k)}_i = \theta^{(k)}_{1-i} = \theta^{(k)}_{2n+1-i}.
$$
Similarly,
$$
\begin{aligned}
  \sum_{i=1}^n
  \theta^{(k)}_i
  \left(
    m_{j - i} + m_{i+j-1}
  \right)
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j-l-\frac12)k\pi}{n}.
\end{aligned}
$$
%
So
$$
\begin{aligned}
\sum_{i = 1}^n \theta^{(k)}_i \, w_{ij}
&=
- \theta^{(k)}_j \, m_0
+\sum_{l=0}^{b} m_l \, 2 \, \cos\frac{lk\pi}{n} \, \cos\frac{(j-\frac12)k\pi}{n}
\\
&= \theta^{(k)}_j \, \lambda_k,
\end{aligned}
$$
with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue}.
}
%
\begin{equation}
\lambda_k = m_0 + \sum_{l = 1}^b 2 \, m_l \cos \frac{k l \pi}{n}.
\label{eq:wband_eigenvalue}
\end{equation}

\section{Numerical results}

1. Single-bin update.
Show the error versus $\lambda$.

2. Multiple-bin update scheme.
